An Integral Projection Modeling Approach to Demographic Consequences of Multiple Partner Mutualisms

### 2023-06-08 Manuscript/Manuscript_Am_Nat.tex
Update Manuscript_Am_Nat.tex
I started commenting, then just started a new draft

  Mutualisms are species interactions where all participants benefit, leading to higher individual fitness and increased population growth rates. 
They are among the most widespread species interactions \cite{Chamberlain2014, BoucherDouglasH.1985} with diverse and dynamic consequences \cite{Bronstein1994,Chamberlain2014,Frederickson2013}, often deteriorating into other types of interactions, including parasitism \cite{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022}.
One reason they are so ubiquitous is they take many forms in nature, including defense for food or housing \cite{Willmer1997}, pollination and dispersal for food \cite{Sakai2002,Burns2004}, resource uptake for housing \cite{Holland2010}, etc. 
Mutualisms are typically considered more context dependent than other species interactions \cite{Chamberlain2014,Frederickson2013}, meaning strength of the interaction is often determined by the surrounding environment and species.

  Historically, pairwise studies, where two species engage in mutualism with only each other, were central in the field of mutualisms, it has since been shown that pairwise studies are poor predictors of the effects of multi-species mutualistic interactions \cite{Afkhami2014,Palmer2010}, where more than two species engage in mutualisms.
The differences in mutualistic outcomes are due to differences in partner quality \cite{Bascompte2019,Stanton2013,Frederickson2013,Jones2015, Ness2006}, differences in the types of benefits offered \cite{Kiers2003,Afkhami2014}, and direct interactions between partners \cite{Sun2019,Heath2009,Heath2014,Grutter2003}.
For these reasons, pairwise studies cannot be accurately used to predict the outcomes of multi-species mutualisms \cite{Palmer2010, Stanton2013, Chamberlain2014, Song2020}.
Studies on the effects of diverse multi-species mutualism systems are necessary to help us understand the demographic effects of partner diversity in mutualisms \cite{Bascompte2019}. 
Multi-species mutualism partner diversity causes varied demographic effects on the population of the focal mutualist which can be explained by a number of mechanisms.
In some cases, the quality of the benefits offered varies leading to some partners which benefit the individual, some which have neither a positive or negative impact on the individual, and some which negatively impact the host \cite{Bronstein1994,Bronstein2001a,Afkhami2014,Song2020,West2007,Frederickson2013,Jones2015}. 
In cases where the partner offers benefits, the function, or type of benefits, offered by the partner can vary \cite{Stanton2003}.

  This variation in partners can lead to net benefits, no net benefits, or net costs for the focal mutualist.
In this study I focus on the positive effects of partner diversity which can be explained by several mechanisms described below. 
When there is a consistent partner hierarchy based on benefits offered, a more diverse sample of the partner community may be more likely to include the highest quality partner (the one which offers the focal mutualist the most benefit) \cite{Frederickson2013}.
When this leads to the fitness of a focal mutualist interacting with multiple partners being equal to the fitness of a focal mutualist interacting with only the highest quality partner, sampling effect can explain the positive effects of partner diversity \cite{Batstone2018}. 
Partners can also vary functionally, meaning each partner offers a different type of reward to the focal mutualist \cite{Stachowicz2005,Bronstein2006,Stanton2003}.
When this leads to a higher fitness for the focal mutualist than any single interaction, complementarity explains the benefits of partner diversity \cite{Batstone2018}. 
Partners can also have asynchronous responses to the environment, either spatially \cite{Ollerton2006} or temporally \cite{Alarcon2008}.Multiple partners can act as a 'portfolio' leading to more even fitness benefits across environmental or temporal heterogeneity, through the portfolio effect \cite{Batstone2018,Lazaro2022}.

  Partner diversity can have different effects depending on whether partners are present all at once or sequentially (partner turnover) \cite{Djieto-Lordon2005, Ness2006, Bruna2014}. 
Turnover can happen at different timescales \cite{Oliveira1999,Horvitz1986} from minutes to years. 
The degree and order of partner turnover can have important impacts on the fitness of the focal mutualist.
The degree, or rapidity, of partner turnover can impact the level of benefits recieved by the focal mutualist particularly if the benefits continue to accumulate or if they saturate over time \cite{Sachs2004}.
The direction of turnover can also have a significant impact, particularly if the most beneficial partner changes across the ontogeny of the focal mutualists life \cite{Fonseca2003}.
For example, susceptibility to enemies can change across life stages \cite{Boege2005,Barton2010} so the focal mutualist benefits the most when defensive partners align with the more vulnerable life stages of focal mutualists \cite{Djieto-Lordon2005}.
Ant visitation to extra-floral nectar producing plants are a classic and well-studied example of interspecific mutualisms which are often multispecies and have dynamic turnover patterns, making these interactions great systems to study the effects of partner diversity.
A subset of these interactions, defensive ant-plant mutualisms are common interactions in which plants provide food and/or ``housing'' to ants which in turn defend them from herbivores \cite{Bronstein1998, Bronstein2006}. 
While these interactions have been well studied \cite{Ness2006,Beattie1985,Schultheiss2022}, few have considered how diversity within ant defender guilds \cite{Stanton2013} or temporal fluctuations in partner interactions \cite{Trøjelsgaard2015} affect the overall benefits of mutualism for the plant partner.

@@

  Analyzing an IPM requires discretizing the composite IPM into a matrix to calculate the dominant eigenvalue. 
Each component of the IPM (each set of ant partner specific vital rates) can be discretized into its own $ 200 \times 200$ matrix \as shown in figure ***. 
The possibility of eviction, when individuals are predicted to grow outside of the possible size classes, was avoided by adding probabilities of growing smaller or larger than the existing boundaries, as done by many others***.
We used the composite IPM to quantify the effects of partner diversity on the intrinsic growth rate of cholla, $r$ $ln(\lambda)$. 
We calculated the $r$ for each combination of ant partners: complete vacancy; \textit{L. apiculatum} and vacancy; \textit{C. opuntiae} and vacancy; other and vacancy; \textit{L. apiculatum}, \textit{C. opuntiae}, and vacancy, \textit{L. apiculatum}, other, and vacancy; \textit{C. opuntiae}, other, and vacancy; and all ant partners and vacancy.
The relative distributions allowed us to determine if either complementarity or sampling effect were at play in the tree cholla-ant defense system.
Sampling Effect requires that the $r$ when all possible partners are present is equal to the $r$ of the cholla population when only the best partner is present.
Complementarity requires that the $r$ when all possible partners are present is greater than the $r$ of any other combination of partners or any single partner.

>>>>

  Mutualisms are species interactions where all participants benefit, leading to higher individual fitness and increased population growth rates. 
They are among the most widespread species interactions \cite{Chamberlain2014, BoucherDouglasH.1985} with diverse and dynamic consequences \cite{Bronstein1994,Chamberlain2014,Frederickson2013}, often deteriorating into commensalism or parasitism,  \cite{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022}.
%One reason they are so ubiquitous is they take many forms in nature, including defense for food or housing \cite{Willmer1997}, pollination and dispersal for food \cite{Sakai2002,Burns2004}, resource uptake for housing \cite{Holland2010}, etc. 
Mutualisms are considered more context dependent than other species interactions \cite{Chamberlain2014,Frederickson2013}, meaning the magnitude and sign of interaction strength are often determined by the surrounding environment and species' identities.

  Historically, pairwise studies, where two species engage in mutualism with only each other, were central in the field of mutualisms, it has since been shown that pairwise studies are poor predictors of the effects of \tom{multi-species mutualistic interactions}{It is very important to give this an operational definition early in the paper.} \cite{Afkhami2014,Palmer2010}, where more than two species engage in mutualisms.
The differences in mutualistic outcomes are due to differences in partner quality \cite{Bascompte2019,Stanton2013,Frederickson2013,Jones2015, Ness2006}, differences in the types of benefits offered \cite{Kiers2003,Afkhami2014}, and direct interactions between partners \cite{Sun2019,Heath2009,Heath2014,Grutter2003}.
\tom{For these reasons, pairwise studies cannot be accurately used to predict the outcomes of multi-species mutualisms \cite{Palmer2010, Stanton2013, Chamberlain2014, Song2020}.}{This re-states something you said two sentences ago.}
Studies on the effects of diverse multi-species mutualism systems are necessary to help us understand the demographic effects of partner diversity in mutualisms \cite{Bascompte2019}. 
Multi-species mutualism partner diversity causes varied demographic effects on the population of the focal mutualist which can be explained by a number of mechanisms.
In some cases, the quality of the benefits offered varies leading to \tom{some partners which benefit the individual}{what is implicit here is that your definition of multi-species mutualism is where a focal species has multiple partners -- that is not at all clear.}, some which have neither a positive or negative impact on the individual, and some which negatively impact the host \cite{Bronstein1994,Bronstein2001a,Afkhami2014,Song2020,West2007,Frederickson2013,Jones2015}. 
In cases where the partner offers benefits, the function, or type of benefits, offered by the partner can vary \cite{Stanton2003}.\tom{}{The structure of this paragraph is disorienting because you have a list of mechanisms in sentence two, then you introduce new ideas, then you return to these same mechanisms here.}

  \tom{This variation in partners can lead to net benefits, no net benefits, or net costs for the focal mutualist.}{Are there examples of all of these?}
\tom{In this study I focus on the positive effects of partner diversity which can be explained by several mechanisms described below.}{``I'' is not the right pronoun for a multi-autho paper, and the tone of this sentence makes your focus on positive effects seem arbitrary and not well supported.} 
When there is a consistent partner hierarchy based on benefits offered, a more diverse sample of the partner community may be more likely to include the highest quality partner (the one which offers the focal mutualist the most benefit) \cite{Frederickson2013}.
When this leads to the fitness of a focal mutualist interacting with multiple partners being equal to the fitness of a focal mutualist interacting with only the highest quality partner, sampling effect can explain the positive effects of partner diversity \cite{Batstone2018}. 
Partners can also vary functionally, meaning each partner offers a different type of reward to the focal mutualist \cite{Stachowicz2005,Bronstein2006,Stanton2003}.
When this leads to a higher fitness for the focal mutualist than any single interaction, complementarity explains the benefits of partner diversity \cite{Batstone2018}. 
Partners can also have asynchronous responses to the environment, either spatially \cite{Ollerton2006} or temporally \cite{Alarcon2008}.Multiple partners can act as a 'portfolio' leading to more even fitness benefits across environmental or temporal heterogeneity, through the portfolio effect \cite{Batstone2018,Lazaro2022}.

  Partner diversity can have different effects depending on whether partners are present all at once or sequentially (partner turnover) \cite{Djieto-Lordon2005, Ness2006, Bruna2014}. 
Turnover can happen at different timescales \cite{Oliveira1999,Horvitz1986}, from minutes to years. 
The degree and order of partner turnover can have important impacts on the fitness of the focal mutualist.
The degree, or rapidity, of partner turnover can impact the level of benefits recieved by the focal mutualist particularly if the benefits continue to accumulate or if they saturate over time \cite{Sachs2004}.
The direction of turnover can also have a significant impact, particularly if the most beneficial partner changes across the ontogeny of the focal mutualist \cite{Fonseca2003}.
For example, susceptibility to enemies can change across life stages \cite{Boege2005,Barton2010} so the focal mutualist benefits the most when defensive partners align with the more vulnerable life stages of focal mutualists \cite{Djieto-Lordon2005}.
Ant visitation to extra-floral nectar producing plants are a classic and well-studied example of interspecific mutualisms which are often multispecies and have dynamic turnover patterns, making these interactions great systems to study the effects of partner diversity.
A subset of these interactions, defensive ant-plant mutualisms are common interactions in which plants provide food and/or ``housing'' to ants which in turn defend them from herbivores \cite{Bronstein1998, Bronstein2006}. 
While these interactions have been well studied \cite{Ness2006,Beattie1985,Schultheiss2022}, few have considered how diversity within ant defender guilds \cite{Stanton2013} or temporal fluctuations in partner interactions \cite{Trøjelsgaard2015} affect the overall benefits of mutualism for the plant partner.

@@

  Analyzing an IPM requires discretizing the composite IPM into a matrix to calculate the dominant eigenvalue. 
Each component of the IPM (each set of ant partner specific vital rates) can be discretized into its own $ 200 \times 200$ matrix as shown in figure ***. 
The possibility of eviction, when individuals are predicted to grow outside of the possible size classes, was avoided by adding probabilities of growing smaller or larger than the existing boundaries, as done by many others***.
We used the composite IPM to quantify the effects of partner diversity on the intrinsic growth rate of cholla, $r$ $ln(\lambda)$. 
We calculated the $r$ for each combination of ant partners: complete vacancy; \textit{L. apiculatum} and vacancy; \textit{C. opuntiae} and vacancy; other and vacancy; \textit{L. apiculatum}, \textit{C. opuntiae}, and vacancy, \textit{L. apiculatum}, other, and vacancy; \textit{C. opuntiae}, other, and vacancy; and all ant partners and vacancy.
The relative distributions allowed us to determine if either complementarity or sampling effect were at play in the tree cholla-ant defense system.
Sampling Effect requires that the $r$ when all possible partners are present is equal to the $r$ of the cholla population when only the best partner is present.
Complementarity requires that the $r$ when all possible partners are present is greater than the $r$ of any other combination of partners or any single partner.
